The role of standing variation in parallel adaptation

### 2014-09-03 standing-variation-paper.tex
wrote intro

In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive allele, such that parallel mutation must be due to multiple mutations after the environmental shift. In that setting we found that there was a characteristic geographic scale, over which we would expect multiple instances of our adaptive allele to have arisen in parallel. This characteristic length could be expressed in terms of a simple compound parameter determined by our parameters of interest.
Recall questions of origin of adaptations

References on parallel adaptation, standing variation, etc.:
  coop,
  pennings and hermisson.
Note pennings and hermisson found standing deleterious variation was very important.

@@

@@

@@

\subsection{Standing Variation -- Results} 
The quantity that is perhaps easiest to compute is the mean time until adaptation.
Fix some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

@@

In Figure \ref{G6PD_standing_var_proportion}

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
\caption{ %
{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the sickle-cell example.  XXX say more.
} \label{G6PD_standing_var_proportion}
\end{center}
\end{figure}

\marginnote{do mean (and variance?) of distance between haplotypes?}

influence of migration on things -- plot these versus migration

>>>>

There is increasing number of examples of different populations within species often
adapt to shared environments via independent genetic changes. 
In some cases this convergent evolution leads to differing phenotypes
in response to shared selection pressures
\citep{} in other cases the degree of convergence is precise down the
same nucleutide change. Such convergent evoltuion within populations
has been seen for many
carefully studied phenotypes in a range of species, including drug resistance, pathogen resistance, insecticideresistance, and pigmentation. In many of these examples we 
need to be careful about whether they changes are truely convergent or
whether selection pressures differ subtly. 
We note that we treat the phrases convergent and parallel evolution
synonymously, as we are concerned with adaptation within a single
species that can occur via a different or shared genetic routes \citep{Arendt:08}.

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. These include the extent
to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{karasov2010}, and whether
species should be regarded to cohesive units. Convergence within
populations also effects our
ability to detect adaptation from population genomic data as no one
allele sweeps fixation over the entire area affected by a selection
pressure \citep{softsweepsIII}.

Convergent evolution can occur even within a well mixed population
subject to a constant selection pressure, either through selection on
multiple mutations present as standing variation within population
before selection pressures switch \citep{Orr:01,softsweepsI} or due to
multiple adaptive alleles arising after selection pressures switch \citet{softsweepsII}. 
This previous work has shown that 
a primary determinant of the probability that multiple alleles contribute to adaptation
is the population scaled mutation rate \citep[see ][ for a review]{MesserPetrov}.

Geographically limited dispersal also increases the chance of
convergent evolution. For example when selection pressures are
geographically patchy this can increasing the probability of
convergent evolution as alleles are unable to spread through intervening populations \citep{RalphCoop:14}.
In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive
allele, such that parallel mutation must be due to multiple mutations
after the environmental shift. In that setting we found that there was
a characteristic geographic scale, over which we would expect multiple
instances of our adaptive allele to have arisen in parallel. This
characteristic length could be expressed in terms of a simple compound
parameter determined by our parameters of interest.

Here we extend our spatial model from \citep{RalphCoop} to include standing
variation present at mutation-selection balance before selection
pressures switch.

@@

In Figure \ref{Fig-G6PD-charlength} we show the range of characteristic lengths
as a function 
calculated for our G6pd parameters.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
\caption{ %
{\bf Characteristic length,} as a function of selective advantage,
holding the prior disadvantage of the allele fixed $s_d=0.05$. 
We compare this to the corresponding quantity without standing variation (e.g.\ from \cite{ralphcoop2010}) and to the quantity only considering standing variants (e.g.\ $\sqrt{\lambda_0}$).
} \label{Fig-G6PD-charlength-sb}
\end{center}
\end{figure}

@@

\subsection{Standing Variation -- Results} 
The quantity that is perhaps easiest to compute is the mean time until adaptation.
If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

@@

In Figure \ref{G6PD_standing_var_proportion} we show the proportion of
alleles that spread from standing variation, and the proportion of
geographic space covered by standing variants.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_standing_var_proportion}
\caption{ %
{\bf Mean proportion of patches arising from standing variation.} The parameters given are for the G6PD example.  XXX say more.
} \label{G6PD_standing_var_proportion}
\end{center}
\end{figure}
